---
title: "A Story of Revisions to U.S. Employment"
author: "Chris Grieves"
format: revealjs
editor: visual
---

```{r, include=F}
#load libraries
library(tidyverse)
library(scales)
library(ggforce)
#pulled from here
#https://www.bls.gov/cew/revisions/qcew-revisions.csv
#get revisions to the QCEW
qcew_revisions <- read_csv("qcew-revisions.csv")

#read in the CES benchmark revisions preliminary and final
ces_revisions <- read_csv("ces_revisions.csv")

#add the date and the difference between the prelim and final
ces_revisions <- ces_revisions %>% 
                  mutate(diff=(Final-Prelim)*1000,
                         date = as.Date(paste0(March,"-3-1"),
                          format="%Y-%m-%d") ) 

#fix the names of the dataset so it's easier to use
names(qcew_revisions) <- tolower(make.names(names(qcew_revisions))) 

#I am only considering US revisions right now
qcew_revisions <- qcew_revisions %>% 
 #Subset data to national employment and drop establishment counts
                    filter(area=="United States",                
                           str_detect(field,"Employment")) %>%
#convert 'field' into a month name and number field
               mutate(monthn = trimws(str_remove(field," Employment")),
                           month = map_int(monthn,\(x){ which(x==month.name)})) %>% 
                    #CES only keep the initial QCEW values and its two revisions
                    select( year, quarter, initial.value,monthn, month,
                            first.revised.value) %>% 
                    #rename the values so that they're easier to use
                    rename( cut1 = initial.value, 
                            cut2 = first.revised.value) %>% 
                    #convert columns to numeric
                    #also create a date variable
                    mutate(across(matches("cut"),\(x){as.numeric(x)}),
                           date= as.Date(paste0(year,"-",month,"-1"),
                                         format="%Y-%m-%d"))
```

# Disclaimer: Thoughts and opinions are my own and are not those of the Bureau of Labor Statistics

## U.S. Employment

U.S. Employment is measured by the Bureau of Labor Statistics (BLS) in two ways:

-   The Current Employment Statistics (CES) Survey
    -   Monthly employment survey asking a sample of employers "How many people do you have on your payroll?"
-   The Quarterly Census of Employment and Wages (QCEW)
    -   Large administrative program that collects employment and wages via each state's Unemployment Insurance (UI) program

## Why two measurements?

-   QCEW is a **census**
    -   **census** - complete count of a population
    -   Problem is that it is on a 4 month lag.
    -   More realistically it is an 8 month lag.
-   CES is produced monthly by a survey
    -   **survey** - using random sampling to gather information about a population.
    -   Only one month lag! - gets users information faster
    -   Subject to sampling error

## Who uses this data?

-   The Federal Reserve relies on this information to set interest rate and make other important economic decisions
-   States use it to make decisions about their own policies

## QCEW Revisions

- QCEW data for a given quarter released ~5 months after that quarter 
- Every release previous quarters data are updated
- We will call each months subsequent releases cut 1, cut 2, and so on..
- We call the difference between cuts, the QCEW revision
-   QCEW Revisions can be found on the BLS website [here](https://www.bls.gov/cew/revisions/)
-   We will care about cut 1 and cut 2 in this presentation


## Benchmarking

-   CES **Benchmarks** to QCEW for the month of March of the previous year.
    -   **Benchmarking** is the process of setting a value to a perceived more credible value.
-   CES releases two benchmark revisions, i.e. "How far away from the 'truth' am I?":
    -   Preliminary Benchmark in August using cut 1 of the QCEW
    -   A Final Benchmark in February using cut 2 of the QCEW
-   My question: How do QCEW revisions affect the differences between the Preliminary and Final Benchmark CES revision?


## CES Revisions

-   Final Revisions on BLS website [here](https://www.bls.gov/web/empsit/cestn.htm#Revisions)
-   Preliminary Revisions
    -   Current year's can be found [here](https://www.bls.gov/web/empsit/cesprelbmk.htm)
    -   Also needed to gather old revisions using the [Wayback Machine](http://wayback.archive.org/) to find archived Preliminary Benchmark announcements since they overwrite them each year.

## CES Revisions

```{r echo=FALSE}

ces_revisions %>% 
  select(-March,-diff) %>% 
  pivot_longer(c("Prelim","Final")) %>% 
  mutate(name=factor(name,levels=c("Prelim","Final"))) %>% 
  ggplot(aes(x=date,y=value,fill=name)) +
  geom_bar(stat="identity",position="dodge") +
  scale_y_continuous(n.breaks=6)+
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  labs(title="CES Prelim and Final Benchmark Revisions (in thousands)",
       y="Revision (in thousands)",
       x="Year",
       fill="Revision") + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

# A Quantity of interest would be to look at the difference Final Revision - Preliminary Revision

## Final - Prelim

```{r echo=F}
ces_revisions %>% 
  ggplot(aes(x=date,y=diff/1000)) +
  geom_bar(stat="identity",fill="blue") + 
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  labs(title="Final Minus Prelim Benchmark Revisions",
       y="Difference (in thousands)",
       x="Year") + 
  theme(plot.title = element_text(hjust = 0.5)) 
```
## Why am I interested in this difference?

- If CES uses cut 1 of QCEW for the Preliminary Benchmark Revision and cut 2 for the Final Benchmark Revision, we should expect this difference to correlate with QCEW revisions.


# Now Let's Consider QCEW

## QCEW Levels Look Like This

```{r, echo=F}
###cut1 cut2 plot
qcew_revisions %>% 
  pivot_longer(matches("cut")) %>% 
  ggplot(aes(x=date,y=value,col=name)) + 
  geom_line() +
  scale_y_continuous( labels= scales::comma_format(),
                      n.breaks = 8 ) + 
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y") +
  labs(title="Initial QCEW Employment (Cut 1) vs First Revised (Cut 2)",
       col="Cut",
       y= "Employment",
       x= "Date")+ 
  theme(plot.title = element_text(hjust = 0.5)) 

```

# OK, those lines look really close, why should I care?

## If we look at the difference between cut 2 and cut 1, we get...

```{r echo=F}
#plot revisions BUT only highlight march
qcew_revisions %>%  
  mutate(rev=cut2-cut1) %>% 
  ggplot(aes(x=date,y=rev)) + 
  geom_line(alpha=.5) + 
  geom_hline(yintercept = 0,col=alpha("black",alpha = .5),linetype = 2)+
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  scale_y_continuous( labels= scales::comma_format()) + 
  labs(title="QCEW Revisions Cut 1 to Cut 2",
       y= "Revision",
       x= "Date")+ 
  theme(plot.title = element_text(hjust = 0.5)) 

```

## Revisions are strictly positive starting in 2021

```{r echo=F}
#plot revisions BUT only highlight march
qcew_revisions %>%  
  mutate(rev=cut2-cut1) %>% 
  ggplot(aes(x=date,y=rev)) + 
  geom_line(alpha=.5) + 
  geom_hline(yintercept = 0,col=alpha("black",alpha = .5),linetype = 2)+
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  scale_y_continuous( labels= scales::comma_format()) + 
  labs(title="QCEW Revisions Cut 1 to Cut 2",
       y= "Revision",
       x= "Date")+ 
  theme(plot.title = element_text(hjust = 0.5)) 
```

## 

- If I overlay the difference between Prelim and Final CES revision
- [red]{style="color:red;"} points are CES differences. **Black** just to highlight March.

```{r, echo=FALSE}
#plot revisions BUT only highlight march
cesplot <- qcew_revisions %>%  
  mutate(rev=cut2-cut1) %>% 
  ggplot(aes(x=date,y=rev)) + 
  geom_line(alpha=.5) + 
  geom_point(aes(x=date,y=rev,shape=if_else(month==3,15,NA) ))+
  scale_shape_identity()+
  geom_hline(yintercept = 0,col=alpha("black",alpha = .5),linetype = 2)+
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  scale_y_continuous( labels= scales::comma_format()) + 
  labs(title="QCEW Revisions Cut 1 to Cut 2",
       y= "Revision",
       x= "Date") + 
  geom_point(aes(x=date,y=diff),col="red",size=2,
             data=ces_revisions %>% filter(year(date)>=2017))+ 
  theme(plot.title = element_text(hjust = 0.5)) 
cesplot
```

## 
- We can make a good guess at the change in revision from Prelim to Final for this missing point

```{r, echo=F}
cesplot + geom_segment(aes(x = mdy("9-1-2024"), y = 155e3, xend = mdy("4-1-2024"), yend = 155e3),
                  arrow = arrow(length = unit(0.5, "cm")),col="red")+ 
  theme(plot.title = element_text(hjust = 0.5))

```

## This point is also stands out (June 2022)

```{r echo=F}
#plot revisions BUT only highlight march
fedplot <- qcew_revisions %>%  
  mutate(rev=cut2-cut1) %>% 
  ggplot(aes(x=date,y=rev)) + 
  geom_line(alpha=.5) + 
  geom_hline(yintercept = 0,col=alpha("black",alpha = .5),linetype = 2)+
  scale_x_date(breaks = seq.Date(mdy("1-1-2015"),to = mdy("1-1-2024"),by="year"),date_labels = "%Y")+
  scale_y_continuous( labels= scales::comma_format()) + 
  labs(title="QCEW Revisions Cut 1 to Cut 2",
       y= "Revision",
       x= "Date") + 
  geom_segment(aes(x = mdy("6-1-2021"), y = 625e3, xend = mdy("5-1-2022"), yend = 625e3),
                  arrow = arrow(length = unit(0.5, "cm")),col="red")+ 
  theme(plot.title = element_text(hjust = 0.5))

fedplot

```


## Philadelphia Fed

-   The Philadelphia Fed does quarterly benchmarking to cut 1 of the QCEW

-   In June 2022 they claimed "In the aggregate, 10,500 net new jobs were
added during the period rather than the
1,121,500 jobs estimated by the sum of the
states; the U.S. CES estimated net growth
of 1,047,000 jobs for the period."
- Thats a claim that BLS was off by 1 million jobs!
- [June 2022 Report](https://www.philadelphiafed.org/-/media/frbp/assets/surveys-and-data/benchmark-revisions/2022/early-benchmark-2022-q2-report.pdf)

## This outlier contributed greatly to the error in the fed report 

```{r echo=FALSE}
fedplot
```

## June 2022 Outlier

- Further differences between the Fed's number and CES explained by:
  - subsequent revisions in the QCEW
  - seasonal adjustment
- The outlier itself was caused by employers submitting erroneous "third month zeros" when submitting Unemployment Insurance tax data

## June 2022 Outlier

- No smoking gun found but possibilities include:
  - Calendar effects
    - When do employees get their paychecks and are they getting counted for that month?
  - Payroll provider issues

## Conclusions

-   Administrative data can have its own risks
-   Revisions to the QCEW impact the reliability of the products created by its users

## Presentation Materials

Git repo: https://github.com/kerr14333/employment_revisions 
Slides: https://rpubs.com/kerr14333/employment-revisions
